Student Log-Data from a Randomized Evaluation of Educational Technology: A Causal Case Study

### 2019-10-29 paper.tex
appendix plus

\usepackage{alltt}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}

%\usepackage{endfloat}

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{h_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately 30\% of students as
high-hint users, so we chose $c$ as the
70th percentile of $\bar{h}$,
approximately 0.6.
This definition agreed with the model's classification in
approximately 93\% of cases.

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Oct 28 14:48:44 2019
\begin{table}[ht]
\centering
\begin{tabular}{llrrlrr}
  \hline
  & Effect & Estimate & Std. Error & CI & $T_Z$ & $\rho^2$ \\ 
  \hline
\multirow{4}{*}{Estimate} & ANOVA & -0.18 & 0.05 & [-0.27,-0.08] & 0.00 & 0.00 \\ 
   & ATE & -0.21 & 0.05 & [-0.32,-0.11] & 0.00 & 0.00 \\ 
   & TOT & -0.18 & 0.06 & [-0.30,-0.07] & 0.00 & 0.00 \\ 
   & ANCOVA & -0.19 & 0.05 & [-0.29,-0.10] & 0.00 & 0.00 \\ 
  \hline &  &  &  &  &  &  \\ 
  \multirow{2}{*}{Sensitivity} & [Pretest] & -0.19 &  & [-0.47,0.09] & -9.86 & 0.18 \\ 
   & [Pretest/2] & -0.19 &  & [-0.37,-0.01] & -4.93 & 0.18 \\ 
   \hline
\end{tabular}
\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

\subsection{The Observational Study Approach: Interpretation and Discussion}
The observational study approach to log data seeks to estimate the
effect of a usage parameter, estimated from  computer log data, on
outcomes.
This approach does not exploit randomization---in fact, it discards
the control group entirely.
However, it depends heavily on another feature of RCTs: namely, an
outcome measured for all participants, in this case a posttest.
Causal identification derives not from randomization, but from an
untestable ignorability assumption, coupled with a sensitivity analysis.
In our example, we operationalized strong ignorability with a
propensity-score matching design, but other observational study
designs may apply as well.

We found that hint requests have a negative effect: students who tend
to request hints more often than the median score lower on the
posttest, after controlling for observed covariates.
When hint usage is expressed numerically, we find a monotonically
decreasing relationship between hint requests and posttest scores---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
If, say, weaker students ask for hints more often and score lower
on the posttest regardless of hint usage, posttest scores will be
anticorrelated with hint requests.
This could even be the case after controlling for pretest scores and other
covariates, if these failed to capture all of the relevant dimensions
of student ability.
We found that an unobserved confounder that predicts both hint usage
and posttest scores just as strongly as the pretest could explain the
entire observed relationship.
The data, coupled with the existence of such a missing covariate, are
consistent with both positive and negative effects of hint requests.
On the other hand, covariates that are as important as pretests are
quite rare in practice---this confounding scenario might not be
plausible after all.
A less extreme missing confounder, predicting $Y$ and $H$ with half
the strength of the pretest, cannot explain the observed relationship,
and is only consistent with negative effects of hint requests.

Despite being hobbled by the lack of randomization, and the resultant
threat of confounding, the observational study approach provides weak
evidence that hints are harmful to posttest scores---a potentially
important finding.
This result, if confirmed in other scenarios and other analyses, may
inform student behavior in the tutor and future tutor designs.
However, since it ignores the control group from the CTA1 RCT, it does
not shed light on the CTA1 causal mechanisms---it does not explain the
CTA1 effect.
However, its design and results figure prominently in mediation
analysis, which does.

@@

Since the potential outcome $Y(0,1)$ is undefined in the CTA1 hints
study, any expression (potentially) including $Y(0,1)$ is also
undefined; this includes $\delta(0)$ and $\xi(1)$.
Indeed, since hint behavior is unobserved in the control group
altogether, conventional approaches to mediation analysis do not
apply.
Further, were we to operationalize hint behavior as a continuous
variable, as may seem natural, mediation analysis in this case would
also be nearly impossible.
This is because students assigned to the control group request exactly
zero hints, whereas merely nine members of the treatment group
requested no hints over the course of the study, and these students
barely used the tutor at all.
That being the case, the data provide little to no information about
the distribution of potential outcomes when the software is used
but no hints are requested.

Our solution was to dichotomize hint requests with a median split, as
described above, essentially re-defining the causal question.
Half of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the (treatment group)
median.
To make sense of the respective roles of $H$ and $\hcont$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\hcont$.
In the control group all students are assigned $H=0$ and $\hcont=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\hcont$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\hcont=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the the difference between the average outcome for students in
the treatment group with $H=1$, and the average outcome under the
counterfactual scenario in which they all had $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
\end{equation*}
$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
\end{equation*}
This can be re-written (omitting conditioning on $Z$):
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
\end{align*}
Since $Pr(H=1)=1/2$ ($H$ was defined with a median split),
$\hat{\Delta}=TOT/2$.
Estimated thusly, the indirect effect is negative: $-0.11$, with
a standard error of 0.03.
This analysis suggests that the availability of excessive hints
actually \emph{lowers} the treatment effect.
CTA1 works not because of hints, but despite them.

What would the effect be without hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct effect of $0.16\pm
0.24$.
Including covariates in the regression produces a direct effect of
$0.19\pm 0.21$.

\subsection{Mediation Analysis: Interpretation and Discussion}
Here we have used mediation analysis as a formal framework to
interpret the results of the observational study in Section
\ref{sec:observational} in terms of the overall CTA1 treatment effect.
Doing so requires some conceptual and statistical acrobatics, such as
relying on a dichotomized measurement of hint requests and strictly
counterfactual potential outcomes, but can nevertheless yield
interesting and interpretable results.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.

Like the observational study, these results are subject to
confounding; indeed, the sensitivity analysis in the Section
\ref{sec:observational} suggests that these results, too, may be
explained due to a missing confounder.

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well; modeling hint requests continuously with the $\hcont$ parameter
from (\ref{eq:rasch}) gives potential values $\hcont(1)$ and $\hcont(0)$,
though only $\hcont(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\hcont(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\hcont(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\hcont$ and posttest scores, the
focus of PS is the relationship between $\hcont$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\hcont(1)=r]$,
despite the fact that $\hcont(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\hcont(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\hcont$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\hcont$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\hcont$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

\subsection{Principal Stratification: Interpretation and Discussion}
Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Unlike those two methods, PS understands $\eta(1)$ as a
student-specific baseline characteristic---albeit one that only
manifests in the treatment group.
Students who request more hints need more help, and thus score lower
on the posttest.
On the other hand, the PS analysis suggests that it's these students
who also receive more help---perhaps due to CTA1 hints, or other
features of the curriculum---and hence benefit more from CTA1.

@@

>>>>

\usepackage{alltt}
\usepackage[stable]{footmisc}
\usepackage{fullpage}
\usepackage{setspace}
\usepackage[]{graphicx}
\usepackage[]{color}
\usepackage[OT1]{fontenc}
\usepackage{amsthm,amsmath,amsfonts}
\usepackage{natbib}
\usepackage[colorlinks,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{bm}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{xr}
\usepackage{tikz}
\usepackage{authblk}
\usepackage[section]{placeins}
%\usepackage{endfloat}

@@

The observational study and (especially) the mediation analysis we
describe below require a dichotomous measurement of hint usage: is a
student a high or low hint user?
To this end, we dichotomize $\bar{h}$ by comparing it to a cutoff value, so
that $H_i=\indicator{\bar{h}_i>c}$, where $\indicator{x}=1$ if $x$ is true
and 0 otherwise.
To choose $c$, we fit a two-component modified Rasch mixture model
\citep{rasch} to hint request data, described in an online appendix.
The model classified approximately 30\% of students as
high-hint users, so we chose $c$ as the
70th percentile of $\bar{h}$,
approximately 0.6.
This definition agreed with the model's classification in
approximately 93\% of cases.

@@

% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Mon Oct 28 20:04:28 2019
\begin{table}[ht]
\centering
\begin{tabular}{llrrrrr}
  \hline
  & Effect & Estimate & Std. Error & CI & $T_Z$ & $\rho^2$ \\ 
  \hline
\multirow{4}{*}{Estimate} & ANOVA & -0.18 & 0.05 & [-0.27,-0.08] & 0.00 & 0.00 \\ 
   & ATE & -0.21 & 0.05 & [-0.32,-0.11] & 0.00 & 0.00 \\ 
   & TOT & -0.18 & 0.06 & [-0.30,-0.07] & 0.00 & 0.00 \\ 
   & ANCOVA & -0.19 & 0.05 & [-0.29,-0.10] & 0.00 & 0.00 \\ 
  \hline &  &  &  &  &  &  \\ 
  \multirow{2}{*}{Sensitivity} & [Pretest] & -0.19 &  & [-0.47,0.09] & -9.86 & 0.18 \\ 
   & [Pretest/2] & -0.19 &  & [-0.37,-0.01] & -4.93 & 0.18 \\ 
   \hline
\end{tabular}
\caption{Estimates with and without regression adjustment and sensitivity analysis of the effect of hint usage.} 
\label{tab:matchResults}
\end{table}

@@

@@

Since the potential outcome $Y(0,1)$ is undefined in the CTA1 hints
study, any expression (potentially) including $Y(0,1)$ is also
undefined; this includes $\delta(0)$ and $\xi(1)$.
Indeed, since hint behavior is unobserved in the control group
altogether, conventional approaches to mediation analysis do not
apply.
Further, were we to operationalize hint behavior as a continuous
variable, as may seem natural, mediation analysis in this case would
also be nearly impossible.
This is because students assigned to the control group request exactly
zero hints, whereas merely
5
members of the treatment group
requested no hints over the course of the study, and these students
barely used the tutor at all.
That being the case, the data provide little to no information about
the distribution of potential outcomes when the software is used
but no hints are requested.

Our solution was to dichotomize hint requests, as
described above, essentially re-defining the causal question.
30\% of the treatment group and all of the control group has
$H=0$---requesting hints at a rate less than the
70th percentile of the treatment group.
To make sense of the respective roles of $H$ and $\bar{h}$, the
continuous measure on which it is based, we may imagine a process in
which students are assigned $H$, and under that constraint choose a
value for $\bar{h}$.
In the control group all students are assigned $H=0$ and $\bar{h}=0$, but
in the treatment group half of students have $H=1$ and half have
$H=0$, and $\bar{h}$ varies within those groups.
That structure implies a ``treatment mediator interaction''
where the effect of the mediator on the outcome varies by treatment
status.
Of course, in our example treatment mediator interactions are not
identified, since $H=0$ and $\bar{h}=0$ in the control group, and very
few treatment students asked for 0 hints.
In any event, dichotomizing hint requests likely undersells the role
of hints, but allows a tractable mediation analysis.

@@

This estimate is a close cousin of the ``TOT'' estimate in Table
\ref{tab:matchResults} from the previous section.
Recall that the TOT effect is the average effect of $H$ on treatment
group subjects with $H=1$.
Both estimate contrast outcomes between $H=0$ and $H=1$, and both use
data from only the treatment group.
TOT is the the difference between the average outcome for students in
the treatment group with $H=1$, and the average outcome under the
counterfactual scenario in which they all had $H=0$.
That is,
\begin{equation*}
TOT=\EE[Y(1,1)|Z=1,H=1]-\EE[Y(1,0)|Z=1,H=1]
\end{equation*}
$\hat{\Delta}$ estimates the difference between the average outcome
for students in the entire treatment group, and the average outcome
for the treatment group were $H=0$ for all students.
That is
\begin{equation*}
\Delta=\EE[Y(1)|Z=1]-\EE[Y(1,0)|Z=1]
\end{equation*}
This can be re-written (omitting conditioning on $Z$):
\begin{align*}
\begin{split}
\Delta&=\EE[Y(1,1)|H=1]Pr(H=1)+\EE[Y(1,0)|H=0]Pr(H=0)\\
&\qquad -\EE[Y(1,0)|H=1]Pr(H=1)-\EE[Y(1,0)|H=0]Pr(H=0)
\end{split}
\\[2ex]
&=(\EE[Y(1,1)|H=1]-\EE[Y(1,0)|H=1])Pr(H=1)
\end{align*}
Since $Pr(H=1)=$0.7,
$\hat{\Delta}=$0.7$TOT$.
Estimated thusly, the indirect effect is negative: -0.1, with
a standard error of 0.04.
This analysis suggests that the availability of excessive hints
actually \emph{lowers} the treatment effect.
CTA1 works not because of hints, but despite them.

What would the effect be without hints?
To answer that question, we estimate ``direct'' effects,
$\Xi=\EE[\xi(1)]=\EE[Y(1,H(0))]-\EE[Y(0,H(0))]$.
Like indirect effects, this expression includes the average potential
outcome $\EE[Y(1,H(0))]$; unlike indirect effects, includes a term for
potential outcomes under the control condition.
Since the main contrast of $\Xi$ is due to treatment assignment,
construction of the estimator must account for the randomization
scheme.
This may be done with a regression estimator.
One option is to construct outcome $\tilde{Y}$ as, simply, $Y$ for
subjects in the control group and $\hat{Y}(1,0)$ for subjects in the
treatment group.
Then, regress it on a dummy variable for treatment group and dummy
variables for randomization block.
The clustering of subjects within schools can be accounted for in a
number of ways, such as including random intercepts for school or
computing cluster-robust standard error estimates.
Following the latter route, and using the \texttt{clubSandwich}
package in \texttt{R} \citep{clubsandwich}, we estimated a direct
effect of
0.14$\pm$0.23.
Including covariates in the regression produces a direct effect of
0.18$\pm$0.19.

\section{Principal Stratification}\label{sec:principalStratification}
The observational study approach and mediation analysis both rely on
an ignorability assumption to estimate the effect of $H$ on $Y$, in
the latter case framing this effect as one component of the overall
treatment effect.
Principal stratification \citep[PS;][]{frangakis} takes a different approach entirely.
In the PS approach, $Y(0)=Y(Z=0)$ and $Y(1)=Y(Z=1)$ are the only potential
outcomes for $Y$.
As in mediation analysis, hint behavior has potential outcomes as
well.
Specifically, the rate at which as student requests hints, $\bar{h}$,
has potential values $\bar{h}(1)$ and $\bar{h}(0)$,
though only $\bar{h}(1)$ is relevant in the CTA1 study.

The goal of principal stratification is to estimate ``principal
effects'':
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\bar{h}(1)=r]
\end{equation*}
This is the treatment effect for subjects who \emph{would} request
hints at level $\bar{h}(1)=r$, if assigned to the treatment condition.
Whereas the focus of the observational study and mediation
analysis was the relationship between $\bar{h}$ and posttest scores, the
focus of PS is the relationship between $\bar{h}$ and treatment effects.
Estimating $\tau(r)$ requires estimating $\EE[Y(0)|\bar{h}(1)=r]$,
despite the fact that $\bar{h}(1)$ is never observed at the same time as
$Y(0)$. Note that despite this fact, $\tau(r)$ does not depend on
strictly counterfactual quantities, as in mediation analysis.
$\bar{h}(1)$ is defined for subjects in both randomized groups, but only
observed in the treatment group.

In previous sections $\bar{h}$ was dichotomized into $H$; such
dichotomization is not necessary here.
However, $\bar{h}$ has some disadvantages as a continuous measure of
hint usage.
For one, since it is essentially a sample mean over a subset of each student's
worked problems, much of its variance is driven by the total number of
problems students worked---in particular, its extreme values belong to
those students who barely used the tutor at all.
Further, it does not account for varying difficulty of the tutor's
problems.
As an alternative, we fit a
modified Rasch model \citep[e.g.][]{rasch},
\begin{equation}\label{eq:rasch}
Pr(h_{ip})=logit^{-1}(\eta_i+\delta_{s[p]})
\end{equation}
where $h_{ip}$ is the event that student $i$ requests a hint on
problem $p$, $\eta_i$ is a student parameter, $\delta_{s[p]}$ is a
section-level parameter for $s[p]$, the section that problem $p$ was
drawn from, and $logit^{-1}(x)=\{1+exp(-x)\}^{-1}$ is the inverse logit
function.
Typically, the student parameter $\eta_i$ measures student ability;
here, instead, it measures a student's proclivity to request a hint.
The typical Rasch model includes a problem-level ``difficulty'' parameter, but in our
dataset many problems were worked by very few students, so problem
level parameters would be hard to estimate.
Instead, we included the section-level parameter $\delta_{s[p]}$,
which measures the extent to which students tend to request hints on
problems in section $s$.
This essentially assumes that all problems in each section are equally
likely to elicit a hint.

We incorporated the model for hint requests, (\ref{eq:rasch}), into
the larger principal stratification model, replacing $\eta$ for
$\bar{h}$.
Then, the principal stratification estimand is
\begin{equation*}
\tau(r)=\EE[Y(1)-Y(0)|\eta(1)=r]
\end{equation*}
A full treatment of this variant of principal stratification, including a
discussion of identification and estimation, may be found in
\citet{aoas}.

@@

\section{Comparing Strategies}\label{sec:synthesis}
The goals, assumptions, and processes of matching, mediation analysis,
and principal stratification, when applied to implementation data from
an EdTech RCT, each differ from each other, but also overlap in
interesting ways.
In the case discussed here, the observational study and mediation
approaches appear to give the opposite conclusion from principal
stratification.
This section will compare and contrast the three methods in terms of
 their necessary identification assumptions, their model fitting
 mechanics, and interpretation of their results.

The fundamental difference between the observational study and
mediation approach, on the one hand, and principal stratification on
the other, is in the role of $H$, the intermediate variable.
An observational study or mediation analysis treats $H$ as a causal
agent, an intervention or exposure that affects $Y$.
In mediation analysis, it is also an (intermediate) outcome, affected
by $Z$.
The approach of principal stratification is quite nearly opposite:
$H$ is not an exposure, that ``happens to'' subjects, but rather its
potential value $H_T$ is a characteristic of the subjects.
Observed $H$ reveals what was there all along.

\subsection{Causal Identification}

Observational studies, and mediation analyses that build on them,
require ignorability assumptions such as Strong Ignorability or
Matched Ignorability.
These may be particularly problematic in our study, since students who
are more likely to request hints are also more likely to struggle with
the material, and hence score lower on the posttest.
We addressed this concern in three ways: first, we focused our attention
on the set of problems in the tutor in which students either requested
a hint or made an error, second, with propensity score matching, and
third with sensitivity analysis.
None of these approaches is bulletproof---they each require further
assumptions about how students request hints.
Eliminating problems in which students showed no signs of struggling
is likely to reduce the association between hint requests and
underlying academic ability, but not eliminate it.
For instance, stronger students may be better able to figure out a
difficult problem without access to hints or error feedback than
weaker students.
Matching depends both on the adequacy of the observed covariates to
capture baseline differences between $H=0$ and $H=1$ students, and the
sensitivity analyses we performed assumed that any un-measured
covariate would predict $H$ no better than pretest, or no better than
a covariate half as important as pretest.
These assumptions are impossible to verify from the data at hand;
instead, they require a keen understanding of student practice within
the tutor.

Mediation analysis builds on the matching design, and also requires an
ignorability assumption; it must
additionally overcome the hurdle of the almost complete lack of
overlap between hint request behavior in the control group (i.e. zero)
and in the treatment group.
Our approach of dichotomizing hint request behavior as $H$ provides a
way forward, at the cost of ignoring information that could be gleaned
from more continuous measures of hint requests.

In contrast, no ignorabilty assumptions are necessary for principal
stratification.
That said, an assumption similar to ignorability may play an important
role for interpretation.
Further, principal stratification estimators require a model for $Y(0)$
conditional on $\eta(1)$, which is unobserved.
On the other hand, model misspecification is, in principle,
detectable---an assumption of a particular family of models is not
``untestable.''
Still, the data will, in general, be unable to distinguish between certain
alternative models for $Y(0)$ conditional on $\eta(1)$, that may in
turn lead to qualitatively different conclusions.
In practice, analysts make the untestable assumption that $Y(0)$ conditional on $\eta(1)$ is
distributed similarly to $Y(1)$ conditional on $\eta(1)$.

An additional contrast of note between matching-based studies and
principal stratification relates to transparency.
However complex the process of devising a match may be, the resulting
design is simple and clear---subjects with $H=1$ are compared against
matches with $H=0$.
This transparency may help researchers assess the plausibility of the
match, and diagnose potential problems with resulting estimates.
Though PS does not depend on ignorability, it does depend on a highly
intricate and parametric modeling structure.
If any part of the model is sufficiently misspecified, the results may
be wrong.
Perhaps worse, \citet{feller2016principal} argued that even well-specified
models can give severely biased estimates.
Extensive model interrogation and checking is necessary for PS
analysis, but the complexity of the PS model makes this process
particularly difficult.
\citet{aoas} gives some examples of potentially fruitful model
checking procedures, which are also carried out in the online supplement.

\subsection{Results and Interpretation}

Our observational study found that hint requests have a negative effect: students who
requested hints more often than the 70th
percentile score lower on the posttest, after controlling for observed covariates---as
far as we can tell, the optimal strategy is to never request a hint.
However, this observed effect may be due to confounding.
Mediation analysis essentially interprets this result in terms of the
overall CTA1 treatment effect.
Specifically, our analysis showed that CTA1 affected students'
posttest scores in (at least) two opposite ways: by allowing hints, it
lowered their posttest scores, but its other mechanisms increased
their scores by an even greater amount.

Like in the observational study and in mediation analysis, the
principal stratification analysis found a negative correlation
between hint requesting and posttest scores.
Unlike those two methods, PS understands $\eta(1)$ as a
student-specific baseline characteristic---albeit one that only
manifests in the treatment group.
Students who request more hints need more help, and thus score lower
on the posttest.
On the other hand, the PS analysis suggests that it's these students
who also receive more help---perhaps due to CTA1 hints, or other
features of the curriculum---and hence benefit more from CTA1.

@@

\newpage
\appendix

\section{Rasch Mixture Model of Hint Usage}

To model hint requests with a mixture model, we modeled each hint
request as in \eqref{eq:rasch}.
Then, instead of modeling $\eta_i$ as normal, we assigned it a
two-component normal mixture model:
\begin{equation*}
\eta\sim
p_0\mathcal{N}(\mu_0,\sigma_0)+(1-p_0)\mathcal{N}(\mu_1,\sigma_1)
\end{equation*}
where we constrained $\mu_0<\mu_1$.

\subsection{Stan Code}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{0.969, 0.969, 0.969}\color{fgcolor}\begin{kframe}
\begin{verbatim}
## data{
## //Sample sizes
##  int<lower=1> N; // total number of obs (long format)
##  int<lower=1> nstud; // # students
##  int<lower=1> nsec; // # sections (problems nested in sections)
## 
## // indices
##  int<lower=1,upper=nstud> studentM[N]; //for each obs, which student?
##  int<lower=1,upper=nsec> section[N]; // for each obs, which section?
## 
## // data data
##  int<lower=0,upper=1> hint[N]; // for each obs, was a hint requested?
## 
## }
## parameters{
## 
##  vector[nstud] eta; // ability parameter
##  real<lower=0,upper=1> p1; // mixing proportion
##  //ordered[2] mu;             // locations of mixture components
##  real<upper=0> mu0; //location of mixture component 0
##  vector<lower=0>[2] sigma;  // scales of mixture components
## 
##  real delta[nsec]; // section difficulty parameter
##  //real<lower=0> sigSec; // scale of difficulty parameters
## 
## }
## transformed parameters{
##  real<lower=0> mu1; // location of mixture component 1
##  mu1= -(1-p1)/p1*mu0;
## }
## model{
##  // logs of mixing proportions to save time
##  real log_p1=log(p1);
##  real log_p0=log(1-p1);
## 
## // linear predictor for logit model
##  real linPred[N];
## 
## // priors
##  sigma ~ lognormal(0, 2);
##  mu0 ~ normal(0, 2);
## // sigSec ~ lognormal(0,2);
##  p1~ beta(2,2); // I'd rather the groups have similar sizes
## 
## // model for section difficulty parameters
##  delta~normal(0,3);//sigSec);
## 
## // Rasch model for hint requests
##  for(i in 1:N)
##   linPred[i]= eta[studentM[i]]-delta[section[i]];
## 
##  hint~bernoulli_logit(linPred);
## 
## 
## // mixture model for ability parameters
##  for(n in 1:nstud)
##   target += log_sum_exp(log_p0+normal_lpdf(eta[n]|mu0,sigma[1]),
## 			log_p1+normal_lpdf(eta[n]|mu1,sigma[2]));
## 
## }
\end{verbatim}
\end{kframe}
\end{knitrout}
%\clearpage
\section{Covariate Balance Before and After Matching}
% latex table generated in R 3.4.4 by xtable 1.8-2 package
% Tue Oct 29 14:14:20 2019
\begin{table}[ht]
\centering
\begin{tabular}{lrrrrrr}
  &\multicolumn{3}{c}{Unmatched}&\multicolumn{3}{c}{Matched}\\
 \hline
 & std.diff & z &  & std.diff & z &  \\ 
  \hline
grade9 & -0.34 & -7.05 & *** & 0.03 & 0.44 &     \\ 
  gradehigher & 0.34 & 7.05 & *** & -0.03 & -0.44 &     \\ 
  raceWhiteAsian & -0.64 & -12.83 & *** & 0.03 & 0.75 &     \\ 
  raceBlackMulti & 0.23 & 4.71 & *** & 0.01 & 0.24 &     \\ 
  raceHispAIAN & 0.51 & 10.35 & *** & -0.05 & -0.98 &     \\ 
  xirt & -0.52 & -10.54 & *** & 0.09 & 1.82 & .   \\ 
  spectypical & -0.16 & -3.32 & *** & 0.03 & 0.48 &     \\ 
  specspeced & 0.34 & 7.07 & *** & -0.10 & -1.44 &     \\ 
  specgifted & -0.14 & -3.03 & **  & 0.06 & 1.16 &     \\ 
  esl0 & -0.37 & -7.70 & *** & 0.04 & 0.63 &     \\ 
  esl1 & 0.37 & 7.70 & *** & -0.04 & -0.63 &     \\ 
  stateCT & 0.58 & 11.81 & *** & -0.00 & 0.00 &     \\ 
  stateKY & -0.36 & -7.51 & *** & -0.00 & 0.00 &     \\ 
  stateLA & -0.08 & -1.68 & .   & -0.00 & 0.00 &     \\ 
  stateMI & -0.36 & -7.51 & *** & 0.00 & 0.00 &     \\ 
  stateTX & 0.43 & 8.77 & *** & 0.00 & 0.00 &     \\ 
   \hline
\end{tabular}
\caption{Covariate balance (standardized differences and Z-scores) before and after the propensity score match.} 
\label{tab:balance}
\end{table}

\begin{figure}[!h]
\centering
\includegraphics[width=0.5\textwidth]{output/balancePlot.pdf}
\caption{Covariate balance (standardized differences) before and after
  the propensity score match.}
\label{fig:balance}
\end{figure}

%\clearpage
\section{Stan Code for PS Model}
